---
title: "PM 566: Final Project"
author: Catherine Le
---

PM 566: FALL 2023

# Downloadable Written Report

Click [here](Written-Report.pdf) to download the written report.

# Brief Description

The question I am exploring in this final project is "What is the correlation between physical activity levels (e.g., reported exercise frequency) and diet habits (e.g., daily consumption of vegetables)?" The intention of this final project is to further explore the stereotype that people who eat healthier also exercise often.

To answer this question, I used the data set "Nutrition, Physical Activity, and Obesity - Behavioral Risk Factor Surveillance System" (BRFSS). BRFSS is an on-going, state-based telephone survey conducted by the CDC and state health departments. The data includes self-reported information on nutrition, behavioral risk factors, and physical activity levels for all 50 U.S. states.

To get the data ready for exploration, I checked for missing values and implausible values (i.e., percentages under 0 and over 100). I then wrangled the data by filtering the data for rows that contained the survey questions "percent of adults who engage in no leisure-time physical activity" and "percent of adults who report consuming vegetables less than one time daily." To explore the data for any patterns, I created histograms and scatterplots.

By creating a scatter plot between the variables "percentage of adults engaging in no leisure time activity" and "percentage of adults consuming vegetables less than one time daily", I found that there was a positive correlation between the two. You can explore my findings further in the interactive graphics down below.

# Interactive Figures

```{r, include=FALSE}
# loading packages
library(tidyverse)
library(ggplot2)
library(dplyr)
library(stringr)
library(plotly)
library(RColorBrewer)

# reading in the data
path <- "~/Desktop/PM566-Final/Data/NPAO_National Survey.csv" 
# dataset = nutrition, physical activity, and obesity
npao_survey <- data.table::fread(path)
```

```{r, echo=FALSE}
# filtering the data to remove the sample size is insufficient values
filtered_npao <- npao_survey %>%
  filter(!grepl("Data not available because sample size is insufficient.",
  Data_Value_Footnote, fixed = TRUE))
```

```{r, include=FALSE}
missing_values <- sum(is.na(filtered_npao$Data_Value))
cat("Number of missing values is", missing_values, "\n")
```

```{r, include=FALSE}
# checking for negative values
negative_values <- filtered_npao$Data_Value < 0
count_negative_values <- sum(negative_values)
cat("Number of negative values is", count_negative_values, "\n")
```

```{r, include=FALSE}
# checking for any values above 100
above_100 <- filtered_npao$Data_Value > 100
count_above_100 <- sum(above_100)
cat("Number of values greater than 100 is", count_above_100, "\n")
```

```{r, include=FALSE}
# filter rows for exercise-related questions
exercise_data <- filtered_npao %>%
  filter(str_detect(Question, "Percent of adults who engage in no leisure-time physical activity"))

# filter rows for vegetable consumption-related questions
vegetable_data <- filtered_npao %>%
  filter(str_detect(Question, "Percent of adults who report consuming vegetables less than one time daily"))
```

```{r, include=FALSE}
exercise_data_aggregated <- exercise_data %>%
  group_by(YearStart, LocationDesc) %>%
  summarize(Data_Value = mean(Data_Value, na.rm = TRUE))

vegetable_data_aggregated <- vegetable_data %>%
  group_by(YearStart, LocationDesc) %>%
  summarize(Data_Value = mean(Data_Value, na.rm = TRUE))
```

```{r, echo=FALSE}
exercise_data_aggregated <- mutate(exercise_data_aggregated, DataType = "Exercise")
vegetable_data_aggregated <- mutate(vegetable_data_aggregated, DataType = "Vegetable")
```

```{r, echo=FALSE}
merged_data <- merge(exercise_data_aggregated, vegetable_data_aggregated, 
                     by = c("YearStart", "LocationDesc"), 
                     suffixes = c("_exercise", "_vegetable"))
```

## Interactive Plot #1

```{r, echo=FALSE}
# first interactive graph
gg <- ggplot(merged_data, aes(x = Data_Value_exercise, y = Data_Value_vegetable)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Association Between Physical Activity and Vegetable Consumption",
       x = "% of Adults Engaging in No Leisure-Time Physical Activity",
       y = "% of Adults Consuming Vegetables < One Time Daily")

# Convert ggplot to plotly
plotly_graph <- ggplotly(gg)
plotly_graph
```

Above is an interactive version of the graph in the Results section of my Written Report. The graph is a scatter plot visualizing the association between the "percentage of adults engaging in no leisure time physical activity" and the "percentage of adults consuming vegetables less than one time daily." ggplotly allows us to hover and see the exact (x.y) value of each point. (The data_value_exercise is the % of adults who do not exercise on a daily basis. The data_value_vegetable is the % of adults who do not consume vegetables on a daily basis)

\
The black line represents a linear regression trend. Its positive slope indicates a correlation, signifying a positive association between the two variables.

## Interactive Plot #2

```{r, echo=FALSE}
merged_data %>%
  plot_ly(x = ~Data_Value_exercise, y = ~Data_Value_vegetable,
          type = 'scatter', mode = 'markers', color = ~LocationDesc,
          hoverinfo = 'text',
          text = ~paste(paste("State: ", LocationDesc, sep=""),
                        paste("Year: ", YearStart, sep=""),
                        paste("Avg % of Adults Who Engage In No Physical Activity: ", Data_Value_exercise, sep=""),
                        paste("Avg % of Adults Who Do Not Eat Vegetables Daily: ", Data_Value_vegetable, sep=""), 
                        sep = "<br>")) %>%
  layout(
    xaxis = list(title = "% of Adults Engaging in No Leisure-Time Physical Activity"),
    yaxis = list(title = "% of Adults Consuming Vegetables < One Time Daily"),
    title = "Association between Vegetable Consumption and Physical Activity Level"
  )
```

This plot is similar to the one above. However, the color of the points correspond to the LocationDesc variable in my dataset (i.e., the U.S. state the survey was conducted in). I also further specified the information that appears when you hover over the points. I specified the state, the year the survey was conducted, and the percentages of survey participants that answered yes to the physical activity level and vegetable consumption survey questions.
